Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-10-01 vertical-thermal-review.Rmd
cleanup
@NidhiVinod , I deleted a few non-critical placeholders for refs (issue #73 ) . We could come back to these later

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. 
The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996]. 
Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017b]. 
Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018a], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions (see section 3.2) and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].

@@

Leaf hydraulic, stomatal, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, because water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`).
In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a]. 
This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$  should decrease $g_s$ of canopy leaves more than understory leaves, particularly when water availability is limited.

@@

In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. 
However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig. `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. 
In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs. `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].  
Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021], *although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [**REFS**]*.
The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

@@

As discussed above, rising temperatures and severe droughts place the canopy trees at particularly elevated risk of mortality. 
Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChangeWidespread2021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011]. 
Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020]. 
While wetter forests can prove quite resilient to such changes [@millerReducedImpactLogging2011], this shift to hotter and drier microclimates can often make forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018]. 
Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [**REFS**, @tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @starkReframingTropicalSavannization2020; @mcdowellPervasiveShiftsForest2020].

Canopy disturbance poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019]. 
Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
If such compositional shifts towards more thermally-adapted species fail to keep up with the pace of warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g., **REFS**, @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy buffering and the potential state changes described above.

@@

## Acknowledgements
Thanks to Norbert Kunert, Camille Piponiot, Peter B. Boucher and Andrew B. Davies for providing figure materials. 
This manuscript benefited from feedback from Eleinis Ávila-Lovera, ... and the ForestGEO Ecosystems & Climate lab at SCBI.
Funding was provided by the Smithsonian Institution.

>>>>

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. 
The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996]. 
Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017b]. 
Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018a], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].

@@

Leaf hydraulic, stomatal, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, because water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO_2 and carbon isotope discrimination discussed in section 3.1 (Table 1).
In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a]. 
This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$  should decrease $g_s$ of canopy leaves more than understory leaves, particularly when water availability is limited.

@@

In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. 
However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig. `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. 
In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs. `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].  
Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021].
The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

@@

As discussed above, rising temperatures and severe droughts place the canopy trees at particularly elevated risk of mortality. 
Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChangeWidespread2021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011]. 
Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020]. 
While wetter forests can prove quite resilient to such changes [@millerReducedImpactLogging2011], this shift to hotter and drier microclimates can often make forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018]. 
Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @starkReframingTropicalSavannization2020; @mcdowellPervasiveShiftsForest2020].

Canopy disturbance poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019]. 
Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
If such compositional shifts towards more thermally-adapted species fail to keep up with the pace of warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g.,  @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy buffering and the potential state changes described above.

@@

## Acknowledgements
Thanks to Norbert Kunert, Camille Piponiot, Peter B. Boucher and Andrew B. Davies for providing figure materials. 
This manuscript benefited from feedback from Eleinis Ávila-Lovera and the ForestGEO Ecosystems & Climate lab at SCBI.
Funding was provided by the Smithsonian Institution.
